---
title: "Cleaning & Wrangling"
format: html
editor: visual
---

```{r}
#| message: false
# load packages
library(dplyr)
library(tidyr)
library(readr)
```

## Read in data
```{r}

catch_original <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")
```

## Explore
```{r}
colnames(catch_original)
head(catch_original)
summary(catch_original)
unique(catch_original$Region)
```

# Clean and reshape data

Shortcut for %>% on Mac Cmd+Shift+M

## Use select function
```{r}
catch_data <- catch_original %>% 
  select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum)
head(catch_data)
catch_data <- catch_original %>% 
  select(-All, -notesRegCode)
head(catch_data)
```

## check column types
```{r}
glimpse(catch_data) 
```

## Replace values in a column
```{r}
# change chinook column to numeric
catch_clean <- catch_data %>%
  mutate(Chinook = as.numeric(Chinook))
```

```{r}
# Find NA values
is.na(catch_clean$Chinook) #not very useful

i <- which(is.na(catch_clean$Chinook))
i

catch_data[i,]

catch_clean <- catch_data %>% 
  mutate(Chinook = if_else(condition = Chinook == "I",
                            true = "1",
                           false = Chinook),
         Chinook = as.numeric(Chinook))
catch_clean[i,]
```

## Reshaping
```{r}
catch_long <- catch_clean %>% 
  pivot_longer(cols = -c(Region, Year),
               names_to = "species",
               values_to = "catch")
head(catch_long)

```

```{r}
catch_wide <- catch_long %>% 
  pivot_wider(names_from = "species",
              values_from = "catch")

```

```{r}
catch_long <- catch_long %>% 
  rename(catch_thousands = catch)
head(catch_long)
```

```{r}
catch_long <- catch_long %>%
    mutate(catch = catch_thousands * 1000)
head(catch_long)

```

## Summary stats using group_by() and summarize()
```{r}
mean_region <- catch_long %>%
    group_by(Region) %>%
    summarize(mean_catch = mean(catch))

head(mean_region)
```

```{r}
# number of rows or each group
n_region <- catch_long %>%
    group_by(Region) %>%
    summarize(n = n())

head(n_region)
```

## Filtering rows using filter()
```{r}
sse_catch <- catch_long %>%
    filter(Region == "SSE")

head(sse_catch)
```

## Filtering > 1000000 fish
```{r}
catch_1M <- catch_long %>%
    filter(catch > 1000000)

head(catch_1M)
```

## Filtering just Chinook
```{r}
catch_ChinookSSE <- catch_long %>%
    filter(species == "Chinook", Region == "SSE")

head(catch_ChinookSSE)
```

## Sorting data using arrange()
```{r}
mean_region <- catch_long %>%
    group_by(Region) %>%
    summarize(mean_catch = mean(catch)) %>%
    arrange(mean_catch) #default is ascending

mean_region <- catch_long %>%
    group_by(Region) %>%
    summarize(mean_catch = mean(catch)) %>%
    arrange(desc(mean_catch)) # change to descending

head(mean_region)
```

## split and combine values in columns using separate() and unite()
```{r}
sites_df <- data.frame(site = c("HAW-101",
                                "HAW-103",
                                "OAH-320",
                                "OAH-219",
                                "MAU-039"))
head(sites_df)

sites_df %>%
    separate(site, c("island", "site_number"), "-") #column to split, new col names, separator
head(sites_df)
```

## Exercise splitting
```{r}
## create `cities_df`
cities_df <- data.frame(city = c("Juneau AK",
                                 "Sitka AK",
                                 "Anchorage AK"))
head(cities_df)
cities_df <- cities_df %>%
    separate(city, c("city", "state_code"), " ") 
head(cities_df)
```

## Unite()
```{r}
dates_df <- data.frame(
    year = c("1930",
             "1930",
             "1930"),
    month = c("12",
              "12",
              "12"),
    day = c("14",
            "15",
            "16")
)
head(dates_df)
dates_df %>%
    unite(date, year, month, day, sep = "-")
```

## all together! 
```{r}
catch_original <- read_csv(url("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1", 
                               method = "libcurl"))

mean_region <- catch_original %>%
  select(-All, -notesRegCode) %>% 
  mutate(Chinook = ifelse(Chinook == "I", 1, Chinook)) %>% 
  mutate(Chinook = as.numeric(Chinook)) %>% 
  pivot_longer(-c(Region, Year), 
               names_to = "species", 
               values_to = "catch") %>%
  mutate(catch = catch*1000) %>% 
  group_by(Region) %>% 
  summarize(mean_catch = mean(catch)) %>% 
  arrange(desc(mean_catch))

head(mean_region)
```

