---
title: "Intro to data viz"
format: html
---

## Setup
```{r}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(janitor) # expedite cleaning and exploring data
library(scales) # scale functions for visualization
library(leaflet) # interactive maps
library(DT) # interactive tables
```

```{r}
escape_raw <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/urn%3Auuid%3Af119a05b-bbe7-4aea-93c6-85434dcb1c5e")
```

## Exploring data
```{r}
#| eval: false

colnames(escape_raw)
glimpse(escape_raw)
range(escape_raw$sampleDate)
unique(escape_raw$Species)
```

## Cleaning data
Cleaning column names
```{r}
escape <- escape_raw %>%
  clean_names()

colnames(escape)

```

Calculate annual escapement by species and sasap regions
```{r}
annual_esc <- escape %>% 
  separate(sample_date, c("year", "month", "day"), sep = "-") %>% 
  mutate(year = as.numeric(year)) %>% 
  group_by(species, sasap_region, year) %>% 
  summarize(escapement = sum(daily_count)) %>%
  filter(species %in% c("Chinook", "Sockeye", "Chum", "Coho", "Pink"))

unique(annual_esc$species)
  
```

## Plotting data
```{r}
ggplot(data = annual_esc, mapping = aes(x = species, y = escapement))+
  geom_col()
```

Making bars blue
```{r}
ggplot(annual_esc,
       aes(x = species, y = escapement)) +
    geom_col(fill = "blue")


```

Adding Regions to the plot
```{r}
ggplot(annual_esc,
       aes(x = species, y = escapement, fill = sasap_region)) +
    geom_col()

```

## Create multiple plots
Subsetting data
```{r}

annual_esc_2000 <- annual_esc %>% 
  filter(year %in% c(2000:2016))
unique(annual_esc_2000$year)

```
 Plotting data
```{r}
ggplot(annual_esc_2000, 
       aes(x = year, y = escapement, color = species)) +
  geom_line() +
  geom_point() +
  facet_wrap(~sasap_region,
             scales = "free_y")

```
 
 Customising plot
```{r}
ggplot(annual_esc_2000, 
       aes(x = year, y = escapement, color = species)) +
  geom_line() +
  geom_point() +
  facet_wrap(~sasap_region,
             scales = "free_y")+
  labs(title = "Annual Salmon Escapement by Region",
       y = "Escapement")+
  theme_bw()+
  theme(legend.position = "bottom",
          legend.title = element_blank(),
        axis.text.x = element_text(angle = 45,
                                   vjust = 0.5))+
  scale_y_continuous(labels = comma)
```
 
 Saving plot
```{r}
ggsave("plots/annual_salmon_esc_region.jpg", width = 8, height = 6, units = "in")

```
 
## Interactive visualization
Subset data
```{r}
locations <- escape %>%
    distinct(location, latitude, longitude) %>%
    drop_na()

```

Using the 'DT' package for interactive tables
```{r}
datatable(locations)

```

Use the 'leaflet' package to create an interactive map
```{r}
leaflet(locations) %>% 
  addTiles() %>% 
  addMarkers(lng = ~longitude, lat = ~latitude, popup = ~location)

```

Another leaflet map
```{r}
leaflet(locations) %>%
    addWMSTiles(
        "https://www.gebco.net/data_and_products/gebco_web_services/web_map_service/mapserv?request=getmap&service=wms&BBOX=-90,-180,90,360&crs=EPSG:4326&format=image/jpeg&layers=gebco_latest&width=1200&height=600&version=1.3.0",
        layers = 'GEBCO_LATEST',
        attribution = "Imagery reproduced from the GEBCO_2022 Grid, WMS 1.3.0 GetMap, www.gebco.net"
    ) %>%
    addCircleMarkers(
        lng = ~ longitude,
        lat = ~ latitude,
        popup = ~ location,
        radius = 5,
        # set fill properties
        fillColor = "salmon",
        fillOpacity = 1,
        # set stroke properties
        stroke = T,
        weight = 0.5,
        color = "white",
        opacity = 1
    )

```

