---
title: "R Practice: Functions"
format: html
---

```{r}
library(rvest)
library(readr)
library(dplyr)
library(janitor)
```

## Load species table
```{r}
webpage <- rvest::read_html("https://www.pwrc.usgs.gov/BBL/Bander_Portal/login/speclist.php")

tbls <- rvest::html_nodes(webpage, "table") %>% 
    rvest::html_table(fill = TRUE)

species <- tbls[[1]] %>% 
    janitor::clean_names() %>% 
    select(alpha_code, common_name) %>% 
    mutate(alpha_code = tolower(alpha_code))

head(species, 3)
```

## Download data
```{r}
predator_survey <- read_csv("https://arcticdata.io/metacat/d1/mn/v2/object/urn%3Auuid%3Abf4f0c68-e4e9-49ae-8be4-487de8f7d10b")

nest_data <- read_csv("https://arcticdata.io/metacat/d1/mn/v2/object/urn%3Auuid%3A982bd2fc-4edf-4da7-96ef-0d11b853102d")

egg_measures <- read_csv("https://arcticdata.io/metacat/d1/mn/v2/object/urn%3Auuid%3A4b219711-2282-420a-b1d6-1893fe4a74a6")

```

## Look at data
```{r}
colnames(egg_measures)
colnames(nest_data)
colnames(predator_survey)
unique(egg_measures$species)
```

## Code to assign species code into common name
This cpde will take the species table, and add a new column to the predator survey table using join. The predator table will be the left table, and we want a left join to only add species that are in the predator table. 
```{r}
glimpse(predator_survey)
colnames(species)
predator_survey <- left_join(predator_survey, species, join_by(species == alpha_code))
```

## Function to assign species code into common name
This function will take the species table, and add a new column to any data table using join. The data table will be the left table, and we want a left join to only add species that are in the data table. 
```{r}
trans_sp_code <- function(data_table){
  data_table <- left_join(data_table, species, join_by(species == alpha_code))
}
predator_survey <- left_join(predator_survey, species, join_by(species == alpha_code))
```

